From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-09 Pi+/paper/finite.tex
General formatting

To present the group $\Sn$, the primitive operations we are going to use will be adjacent swaps. When dealing with
permutations on an $n + 1$-element set, there are $n$ adjacent transpositions, transposition number $k$ swapping
elements $(\el{k}, \el{k+1})$. Thus, the generating set would be $\Fin[n]$. There are three relations that we're going
to specify for this presentation -- these are the equations that these generators should satisfy. We visualise them as
braid diagrams in~\Cref{fig:coxeter-braid}.

\begin{enumerate}
  \item[\labelcref{fig:coxeter-a}] Swapping the same two elements two times in a row should be the same as doing nothing
  \item[\labelcref{fig:coxeter-b}] When swapping two distinct pairs of elements, it should not matter in which order the
        swapping happens, that is, we can slide the wires freely.
  \item[\labelcref{fig:coxeter-c}] If the next transposition in the sequence is equal to the the first one, the order of
        operations can be reversed.
\end{enumerate}

@@

The result of this function is the sequence \([k + n - 1, k + n - 2, k + n - 3, \ldots, n]\). Since we think of the
number $m$ as the transposition swapping elements $(\el{m}, \el{m+1})$, the role of this function is to produce a
sequence of transpositions which moves element $\el{k + n}$ left by $k$ places, shifting all the elements in between one
place right (see~\cref{fig:downf}). Then, the directed relation $\longcox$ is defined with the following generators,
inlining the congruence closure in $\longcox$, allowing arbitrary reductions inside the list.

\begin{definition}[$\longcox$]
  \begin{align*}
    \longcancel
     & : \forall n, l, r \to (l \append n \cons n \cons r) \longcox (l \append r)                                                                      \\
    \longswap
     & : \forall n, k, l, r \to (\suc[k] < n) \to (l \append n \cons k \cons r) \longcox (l \append k \cons n \append r)                               \\
    \longbraid
     & : \forall n, k, l, r \to (l \append (n \downf 2 + k) \append (1 + k + n) \cons r) \longcox (l \append (k + n) \cons (n \downf 2 + k) \append r)
  \end{align*}
\end{definition}

@@

\begin{toappendix}
  \begin{definition}[$\longcoxstar$]
    \label{def:longcox-star}
    \begin{align*}
      \reflr{\longcox}
       & : \forall w \to w \longcoxstar w                                                                                  \\
      \transr{\longcox}
       & : \forall w_{1}, w_{2}, w_{3} \to  w_{1} \longcox w_{2} \to w_{2} \longcoxstar w_{3} \to w_{1} \longcoxstar w_{3}
    \end{align*}
  \end{definition}

  \begin{definition}[$\longcoxplus$]
    \label{def:longcox-plus}
    \begin{align*}
      \relr{\longcoxplus}
       & : \forall w_{1}, w_{2} \to w_{1} \longcox w_{2} \to w_{1} \longcoxplus w_{2}                                      \\
      \transr{\longcoxplus}
       & : \forall w_{1}, w_{2}, w_{3} \to  w_{1} \longcox w_{2} \to w_{2} \longcoxplus w_{3} \to w_{1} \longcoxplus w_{3}
    \end{align*}
  \end{definition}
\end{toappendix}

@@

Despite the increased complexity of the generators, the rewriting system $(\List[\Fin[n]],\longcox)$ has the properties
we desire. It satisfies local confluence, that is, the Church-Rosser (diamond) property, and it is terminating, so, by
Newman's lemma, it produces a unique normal form. We follow the terminology
of~\cite*{huetConfluentReductionsAbstract1980a,krausCoherenceWellFoundednessTaming2020} to state our results formally.

@@

From knowing just the inversion counts for all the elements, one can reconstruct the starting permutation. Also, observe
that $\invcount{i} < i$, thus fitting in the $i$-th place of a Lehmer code tuple. As an example, consider the following
tabulated presentation of a permutation of $\Fin[5]$.

\[
  \sigma =
  \begin{pmatrix}
    0      & 1      & 2      & 3      & 4      \\
    \el{2} & \el{1} & \el{4} & \el{0} & \el{3} \\
  \end{pmatrix}
\]

>>>>

To present the group $\Sn$, the primitive operations we are going to use will be adjacent swaps. When dealing with
permutations on an $n + 1$-element set, there are $n$ adjacent transpositions, transposition number $k$ swapping
elements at indices $k$ and $k+1$. Thus, the generating set would be $\Fin[n]$. There are three relations that we're going
to specify for this presentation -- these are the equations that these generators should satisfy. We visualise them as
braid diagrams in~\Cref{fig:coxeter-braid}.

\begin{enumerate}
  \item[\labelcref{fig:coxeter-a}] Swapping the same two elements two times in a row should be the same as doing nothing
  \item[\labelcref{fig:coxeter-b}] When swapping two distinct pairs of elements, it should not matter in which order the
        swapping happens, that is, we can slide the wires freely.
  \item[\labelcref{fig:coxeter-c}] \jk{If the next transposition in the sequence is equal to the the first one, the order of
        operations can be reversed}.
\end{enumerate}

@@

The result of this function is the sequence \([k + n - 1, k + n - 2, k + n - 3, \ldots, n]\). The role of this function
is to produce a sequence of transpositions which moves element at index $k + n$ left by $k$ places, shifting all the
elements in between one place right (see~\cref{fig:downf}). Then, the directed relation $\longcox$ is defined with the
following generators, inlining the congruence closure in $\longcox$, allowing arbitrary reductions inside the list.

\begin{definition}[$\longcox : {\List[\Fin[n]]} \to {\List[\Fin[n]]} \to {\UU}$]
  \begin{align*}
    \longcancel
     & : \forall n, l, r \to (l \append n \cons n \cons r) \longcox (l \append r)                                                                      \\
    \longswap
     & : \forall n, k, l, r \to (\suc[k] < n) \to (l \append n \cons k \cons r) \longcox (l \append k \cons n \append r)                               \\
    \longbraid
     & : \forall n, k, l, r \to (l \append (n \downf 2 + k) \append (1 + k + n) \cons r) \longcox (l \append (k + n) \cons (n \downf 2 + k) \append r)
  \end{align*}
\end{definition}

@@

\begin{toappendix}
  \begin{definition}[$\longcoxstar : {\List[\Fin[n]]} \to {\List[\Fin[n]]} \to {\UU}$]
    \label{def:longcox-star}
    \begin{align*}
      \reflr{\longcox}
       & : \forall w \to w \longcoxstar w                                                                                  \\
      \transr{\longcox}
       & : \forall w_{1}, w_{2}, w_{3} \to  w_{1} \longcox w_{2} \to w_{2} \longcoxstar w_{3} \to w_{1} \longcoxstar w_{3}
    \end{align*}
  \end{definition}

  \begin{definition}[$\longcoxplus : {\List[\Fin[n]]} \to {\List[\Fin[n]]} \to {\UU}$]
    \label{def:longcox-plus}
    \begin{align*}
      \relr{\longcoxplus}
       & : \forall w_{1}, w_{2} \to w_{1} \longcox w_{2} \to w_{1} \longcoxplus w_{2}                                      \\
      \transr{\longcoxplus}
       & : \forall w_{1}, w_{2}, w_{3} \to  w_{1} \longcox w_{2} \to w_{2} \longcoxplus w_{3} \to w_{1} \longcoxplus w_{3}
    \end{align*}
  \end{definition}
\end{toappendix}

@@

Despite the increased complexity of the generators, the rewriting system $(\List[\Fin[n]],\longcox)$ has the properties
we desire. It satisfies local confluence, that is, the Church-Rosser (diamond) property (for example, using
$\longbraid$, we can now show the problematic critical pair \cref{fig:critical-pairs-non-converging} to converge), and it is
terminating, so, Newman's lemma, it produces a unique normal form. We follow the terminology
of~\cite*{huetConfluentReductionsAbstract1980a,krausCoherenceWellFoundednessTaming2020} to state our results formally.

@@

From knowing just the inversion counts for all the elements, one can reconstruct the starting permutation. Also, observe
that $\invcount{i} < i$, thus fitting in the $i$-th place of a Lehmer code tuple. As an example, consider the following
tabulated presentation of a permutation of $\Fin[5]$.
\[
  \sigma =
  \begin{pmatrix}
    0      & 1      & 2      & 3      & 4      \\
    \el{2} & \el{1} & \el{4} & \el{0} & \el{3} \\
  \end{pmatrix}
\]

### 2021-07-09 Pi+/paper/ufin.tex
General formatting

\begin{definition}[$\Fin$]
  \label{def:finite-set}
  The type family $\Fin : \Nat \to \UU$ is the type of finite sets indexed by their cardinality. It is defined
  equivalently in two different ways,
  \begin{gather*}
    \begin{aligned}
      \Fin[n] & \defeq \dsum{k:\Nat}{k < n}
    \end{aligned}
    \qquad\qquad \qquad\qquad
    \begin{aligned}
       & \Fin[0] \defeq \bot                      \\
       & \Fin[\suc[n]] \defeq \top \sqcup \Fin[n]
    \end{aligned}
  \end{gather*}
  Note that $\Fin[n]$ is a set, and we use both definitions interchangeably.
\end{definition}

\begin{definition}[$\isFin$]
  \label{def:isfin}
  We say that a type is finite if it is merely equal to $\Fin[n]$ for some $n$.
  \[
    \isFin[X] \defeq \dsum{n:\Nat}{\SubP{X}{\Fin[n]}}
  \]
  Note that the natural number $n$ need not be truncated, as justified below.
\end{definition}

@@

Since $\BAut[T]$ is a univalent subuniverse, we can characterise its path space. The intuition is that $\BAut[T]$ only
has one point $T_0$, and 1-paths $T_0 \id_{\BAut[T]} T_0$, that is, loops, and higher paths between these loops. The
type of loops on $T_0$, $\loopspace[\BAut[T],T_{0}]$ is shown to be equivalent to $\Aut[T] \defeq T \eqv T$, which is
the group of automorphisms of $T$.

@@

\begin{proposition}
  For any $n, m : \Nat$, and for any types $X, Y, Z$,

  \begin{gather*}
    \begin{aligned}
      \Fin[0]                & \eqv \bot                  \\
      \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m]           \\
      \\
      \bot \sqcup X          & \eqv X                     \\
      X \sqcup \bot          & \eqv X                     \\
      (X \sqcup Y) \sqcup Z  & \eqv X \sqcup (Y \sqcup Z) \\
      X \sqcup Y             & \eqv Y \sqcup Y            \\
      X \times \bot          & \eqv \bot
    \end{aligned}
    \qquad
    \begin{aligned}
      \Fin[1]                & \eqv \top                             \\
      \Fin[n] \times \Fin[m] & \eqv \Fin[n * m]                      \\
      \\
      \top \times X          & \eqv X                                \\
      X \times \top          & \eqv X                                \\
      (X \times Y) \times Z  & \eqv X \times (Y \times Z)            \\
      X \times Y             & \eqv Y \times X                       \\
      X \times (Y \sqcup Z)  & \eqv (X \times Y) \sqcup (X \times Z)
    \end{aligned}
  \end{gather*}
\end{proposition}

>>>>

\begin{definition}[$\Fin$]
  \label{def:finite-set}
  The type family $\Fin : \Nat \to \UU$ is the type of finite sets indexed by their cardinality. It is defined
  equivalently in two different ways,
  \begin{gather*}
    \Fin[n] \defeq \dsum{k:\Nat}{k < n}
    \qquad\qquad \text{or} \qquad\qquad
    \begin{aligned}
       & \Fin[0] \defeq \bot                      \\
       & \Fin[\suc[n]] \defeq \top \sqcup \Fin[n]
    \end{aligned}
  \end{gather*}
  Note that $\Fin[n]$ is a set, and we use both definitions interchangeably.
\end{definition}

\begin{definition}[$\isFin$]
  \label{def:isfin}
  We say that a type is finite if it is merely equal to $\Fin[n]$ for some $n : \Nat$.
  \[
    \isFin[X] \defeq \dsum{n:\Nat}{\SubP{X}{\Fin[n]}}
  \]
  Note that the natural number $n$ need not be truncated, as justified below.
\end{definition}

@@

Since $\BAut[T]$ is a univalent subuniverse, we can characterise its path space. The intuition is that $\BAut[T]$ only
has one point $T_0$, and 1-paths $T_0 \id_{\BAut[T]} T_0$, that is, loops, and higher paths between these loops. The
type of loops on $T_0$, $\loopspace[\BAut[T],T_{0}]$, is shown to be equivalent to $\Aut[T] \defeq T \eqv T$, which is
the group of automorphisms of $T$.

@@

\begin{proposition}
  For any $n, m : \Nat$, and for any types $X, Y, Z$,
  \begin{gather*}
    \begin{aligned}
      \Fin[0]                & \eqv \bot                  \\
      \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m]           \\
      \\
      \bot \sqcup X          & \eqv X                     \\
      X \sqcup \bot          & \eqv X                     \\
      (X \sqcup Y) \sqcup Z  & \eqv X \sqcup (Y \sqcup Z) \\
      X \sqcup Y             & \eqv Y \sqcup Y            \\
      X \times \bot          & \eqv \bot
    \end{aligned}
    \qquad
    \begin{aligned}
      \Fin[1]                & \eqv \top                             \\
      \Fin[n] \times \Fin[m] & \eqv \Fin[n * m]                      \\
      \\
      \top \times X          & \eqv X                                \\
      X \times \top          & \eqv X                                \\
      (X \times Y) \times Z  & \eqv X \times (Y \times Z)            \\
      X \times Y             & \eqv Y \times X                       \\
      X \times (Y \sqcup Z)  & \eqv (X \times Y) \sqcup (X \times Z)
    \end{aligned}
  \end{gather*}
\end{proposition}
